Algebraic Tori in Unitary Groups

### 2014-07-03 Paper.tex
Add some more explicit TODO's and proof sketches

@@

\subsubsection{Algebra's with involutions}

@@

@@

@@

@@

Local conjugacy class statement requires knowing conjugacy classes in the local non-split algebra setting. 
The exact sequence $H^1(F) \rightarrow H^1(Aut(A,F)) \rightarrow H^1(Aut(F))$ tells us these are all conjugate.
A direct argument of this claim would be nice.
(Conjugacy of maximal \'etale sub-algebras of central simple algebras must be known).

@@

\begin{remark}
The cohomological interpretation of this result is as follows.

@@

@@

We may thus define the group schemes $SO_J$ and $SP$:
\[ SO_J(R) =  \{ g\in (M_{n}(C) \otimes_K R)^\times \mid \tau_C(g^t)Jg = J  \} \]
and
\[ SP(R) =  \{ g\in (M_{n}(C) \otimes_K R)^\times \mid \sigma_C(g^t)Jg = J  \}. \]
We readilly observe that $SU_J$ embeds naturally into both of these groups.
\begin{remark}
The group $SP$ is a symplectic group, its isomorphism class does not depend on $J$.
\end{remark}

\begin{remark}
In the special case where $A_0$ and $B$ are split, so that $SU_J$ is the special unitary group of an $E$-valued Hermitian form $H$ associated to $J$ then
$SO_J$ is the orthogonal group associated to the $K$-valued bilinear form $B(z_1,z_2) = \Tr_{E/K}(H(z_1,z_2))$ and the group $SP$ is the symplectic group associated to the symplectic form $S(z_1,z_2) = \Tr_{E/K}(H(yz_1,z_2))$.

Note that given $SO_J$ or $SP$, and the $E$-module structure on the standard representations, the group $SU_J$ may be recovered.
Note also that $U_J = SO_J \cap SP$.
\end{remark}

>>>>

\subsection{Alternating Forms}

As above, but define classical simplectic groups.

Fix a free $K$-module $V$ of dimension $2n$. There is a unique a non-degenerate alternating $K$-valued $K$-bilinear pairing $S$ on $V$.

$\Sp_S = \Sp = \Aut(V,S)$.

The forms of $(V,S)$ are in bijection with $H^1(\GalKbK, \Aut(V,S))$, as there is only a unique form, this cohomology group vanishes.

\subsection{Symplectic Groups}

For the purpose of this paper a symplectic group means a simply connected group of type $C_n$.
The outer automorphism group of such a group is trivial, the center if $\mu_2$.

We may construct the symplectic groups as defined above as follows. Fix a quaternion algebra $A$ over $K$ (we allow $M_2(K)$) together with the standard quaternionic involution $\sigma$. If $A$ is $M_2(K)$ then $\sigma( \left( \begin{smallmatrix} w & x\\ y & z \end{smallmatrix} \right) ) =\left( \begin{smallmatrix}z & {-x}\\ {-y} & w \end{smallmatrix} \right) $

Define then $G$ to be the group scheme whose points over a $K$-algebra $R$ are given by:
\[ G(R) = \{ g\in (M_{n}(A) \otimes_K R)^\times \mid \tau(g^t)g = 1_n \}. \]
We remark the involution $\tau(g^t)$ on $M_{n}(A)$ is of symplectic type.

\begin{theorem}
All forms of $G$ arise this way.
\end{theorem}
\cite[TODO]{book of involutions}

As the outer automorphism group was trivial and the center was $\mu_2$ we obtain the exact sequence:
\[\ldots \rightarrow H^1(\GalKbK,\mu_2) \rightarrow  H^1(\GalKbK,G)  \rightarrow  H^1(\GalKbK, \Aut(G)) \rightarrow  H^2(\GalKbK, \mu_2)  \]
We may thus associate to $G$ a quaternion algebra $A$.
The cohomology set $H^1(\GalKbK,G) $ classifies alternating forms on $A^n$, as above there is a unique such and so this cohomology set is trivial.
\cite[TODO]{book of involutions}.

@@

\subsubsection{Algebras with involutions}

@@

From the above it is clear that to classify the groups $G$ into which $T_{E,\sigma}$ embeds, we must understand which forms may arise as one of
\[ H_{E,\sigma,\lambda} := \tfrac{1}{2}Tr_{E/A}\left(\lambda x\sigma(y)\right) \]
for Hermitian forms or:
\[ B_{E,\sigma,\lambda} := \tfrac{1}{2}Tr_{E/K}\left(\lambda x\sigma(y)\right) \]
for Bilinear forms.
In particular, we must understand the invariants of these forms.

@@

\subsubsection{Algebras with involutions}

We now  explain how the invariants of $\tau_\lambda$ depend on $\lambda$.
Note that in this case the choice of ``base point" is not as canonical and the collection of involutions is really a torsor under the action of the appropriate unit group.

\TODO{fill in details}
\begin{itemize}
\item $\lambda$ choses discriminant of unitary groups.
\subitem This should be intrinsicly clear from definitions... hopefully
\TODO{In the above version I relate the unitary and orthogonal cases... not clear I want to do that for this case until we give construction later}
\item $\lambda$ choses the Clifford invariant and index of orthogonal groups.
\subitem (Same formula as other case)
\item $\lambda$ irrelevent for symplectic groups.
\end{itemize}

@@

\subsubsection{Unitary Groups}

@@

Local conjugacy class statement requires knowing conjugacy classes in the local non-split algebra setting. 
The exact sequence $H^1(F) \rightarrow H^1(Aut(A,F)) \rightarrow H^1(Aut(F))$ tells us these are all conjugate.
\TODO{cite above for direct argument of this claim}

@@

\TODO{cohomological interpretation is supposed to be later I think}
\begin{remark}
The cohomological interpretation of this result is as follows.

@@

\subsubsection{Orthgogonal Groups}

\TODO{copy and paste from that other thing where I wrote the general theorem}
The only meaningful way in which the argument differs from my paper is a semi-ugly case analysis on the clifford algebras.

\TODO{cohomological interpretation is later I think}

\subsubsection{Symplectic Groups}

\TODO{write down the result}

\TODO{cohomological interpretation is later I think}

@@

We may thus define the group schemes $\SO_J$ and $\Sp$:
\[ \SO_J(R) =  \{ g\in (M_{n}(C) \otimes_K R)^\times \mid \tau_C(g^t)Jg = J  \} \]
and
\[ \Sp(R) =  \{ g\in (M_{n}(C) \otimes_K R)^\times \mid \sigma_C(g^t)Jg = J  \}. \]
We readilly observe that $SU_J$ embeds naturally into both of these groups.
\begin{remark}
The group $SP$ is a symplectic group, its isomorphism class does not depend on $J$.
\end{remark}

\begin{remark}
In the special case where $A_0$ and $B$ are split, so that $\SU_J$ is the special unitary group of an $E$-valued Hermitian form $H$ associated to $J$ then
$SO_J$ is the orthogonal group associated to the $K$-valued bilinear form $B(z_1,z_2) = \Tr_{E/K}(H(z_1,z_2))$ and the group $SP$ is the symplectic group associated to the symplectic form $S(z_1,z_2) = \Tr_{E/K}(H(yz_1,z_2))$.

Note that given $\SO_J$ or $\Sp$, and the $E$-module structure on the standard representations, the group $\SU_J$ may be recovered.
Note also that $U_J = \SO_J \cap \Sp$.
\end{remark}

\TODO{prove that this construction is universal - sketch follows}
\begin{itemize}
\item It suffices to prove that rational conjugacy is the same as rational isomorphism.
\item First claim any conjugate embedding (of groups) arises from a conjugate embedding (of algebras with involutions).
\item We want to show that any two embeddings $A \hookrightarrow C$ are conjugate.
        With $A,C$ as defined above but the embedding not necissarily coming from the condition.

        NOTE: we only really care about showing they are $C$ conjugate whenever they are $C\otimes_k \overline{k}$ conjugate.
\item Pretty sure that $A$ is the centralizer in $C$ of the center of $A$.
\item Thus, conjugacy classes of $A$ equivalent to conjugacy classes of center of $A$.
\item Conjugacy classes of the center $Z(A)$ of $A$ are equivalent to $Z(A)$ module structures on the "representation" space of $C$ that are "compatible" with a 
         $[C]$-module structure on the space.

          That is, there is an $[C]$-basis $c_1,\ldots c_n$ of the representation space, such that $Z(A) \cdot c_i \subset [C]\cdot c_i$.
          This should follow from the fact that $Z(A)$ splits $C$.
\item All $[C]$-module structures on the space are equivalent (Hilberts 90), so that any two $c_i$ basis are $C$ equivalent.
\item This reduces us to conjugacy of $Z(A)$ inside $[C]$, which we have previously shown, but is also well known for quaternion algebras.
\end{itemize}

\TODO{Want to prove the following facts [that one hopes are facts]}
\begin{itemize}
\item The normalizer in $\Sp$ (resp. $\SO$) of $\U$ is $\mu_2 \rtimes \U$ (resp. $\U$).  [or something like this]
\subitem  suffices to check over the algebraic closure.
\item The centralizer of the center of $\U$ in $\Sp$ and $\SO$ is $\U$
\subitem suffices to check over the algebraic closure.
\subitem Note that the connected component of the normalizer of the center of $\U$ is the centralizer of the center.
              (first claim reduces to second + claim that normalizer is connected in $\SO$ case)
\end{itemize}

\TODO{Invariants of orthogonal from those of unitary}
\begin{itemize}
\item The discriminant of the unitary involution should determine clifford of the orthogonal.
\item The discriminant of the orthogonal should be determined by the choice of $E$.
\item The index of the unitary and orthogonal should be the same.
\end{itemize}

\TODO{Summary - Given orthogonal which unitary are possible}
\TODO{Summary - Given unitary which orthogonal are possible}

\TODO{cohomological reinterpretation is now or later?}

\subsection{Decomposing Conjugacy Classes of Tori}

\TODO{The questions about which tori are in the image}

\TODO{cohomological reinterpretation is now or later?}
